A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-07-09 tex/sec4-pre-synthesis.tex
[ tex ] Use \nvdash instead of \not\vdash

%\Josh{Throughout this sub-section, we quantify over any bidirectional type system~$(\Sigma, \Omega)$, context $\Gamma : \Cxt_\Sigma(\emptyset)$, raw term $\erase\Gamma \vdash_{\Sigma, \erase\Omega} t$, mode~$d$, and type $A : \Type_\Sigma(\emptyset)$.}

@@

\begin{lemma}\label{thm:Pre?-false}
If\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,\BFalse\,s}$, then $V \not|-_{\Sigma,\Omega} t^\dir{d}$.
\end{lemma}

@@

\begin{proof}
By \cref{thm:generalised-mode-preprocessing}, there is a derivation of $V |-_{\Sigma,\erase\Omega} t^{\dir{d}\,g\,s}$ for some booleans $g$~and~$s$, and then simply check~$g$ and apply either \cref{thm:Pre?-true} or \cref{thm:Pre?-false} to obtain $V |-_{\Sigma,\Omega} t^\dir{d}$ or $V \not|-_{\Sigma,\Omega} t^\dir{d}$.
\end{proof}

%We can also establish the converse of \cref{thm:Pre?-true,thm:Pre?-false}.
%
%\begin{corollary}\label{thm:toPre?-true}
%If\/ $V |-_{\Sigma,\Omega} t^\dir{d}$, then $V |-_{\Sigma,\Omega} t^{\dir{d}\,\BTrue\,s}$ for some boolean~$s$.
%\end{corollary}
%
%\begin{corollary}\label{thm:toPre?-false}
%If\/ $V \not|-_{\Sigma,\Omega} t^\dir{d}$, then $V |-_{\Sigma,\Omega} t^{\dir{d}\,\BFalse\,s}$ for some boolean~$s$.
%\end{corollary}
%
%Whereas it is possible to prove \cref{thm:toPre?-true} directly by induction, converting a mode derivation to a generalised one, the negated antecedent of \cref{thm:toPre?-false} provides little information for constructing a required derivation, and basically we have to construct such a derivation from scratch for a raw term.
%But this is exactly what generalised mode preprocessing does, so we can simply reuse \cref{thm:generalised-mode-preprocessing}.
%And \cref{thm:toPre?-true} can also be proved easily in the same way.
%
%\begin{proof}[Proofs of \cref{thm:toPre?-true,thm:toPre?-false}]
%
%By \cref{thm:generalised-mode-preprocessing}, there is a derivation of $V |-_{\Sigma,\Omega} t^{\dir{d}\,g\,s}$ for some booleans $g$~and~$s$.
%If $V |-_{\Sigma,\Omega} t^\dir{d}$, then $g$~has to be $\BTrue$ since $g = \BFalse$ leads to a contradiction with \cref{thm:Pre?-false}.
%Symmetrically, if $V \not|-_{\Sigma,\Omega} t^\dir{d}$, then $g$~has to be~$\BFalse$ so as not to contradict \cref{thm:Pre?-true}.
%\end{proof}

>>>>

%\Josh{Throughout this sub-section, we quantify over any bidirectional type system~$(\Sigma, \Omega)$, context $\Gamma : \Cxt_\Sigma(\emptyset)$, raw term $\erase\Gamma |-_{\Sigma, \erase\Omega} t$, mode~$d$, and type $A : \Type_\Sigma(\emptyset)$.}

@@

\begin{lemma}\label{thm:Pre?-false}
If\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,\BFalse\,s}$, then $V |/-_{\Sigma,\Omega} t^\dir{d}$.
\end{lemma}

@@

\begin{proof}
By \cref{thm:generalised-mode-preprocessing}, there is a derivation of $V |-_{\Sigma,\erase\Omega} t^{\dir{d}\,g\,s}$ for some booleans $g$~and~$s$, and then simply check~$g$ and apply either \cref{thm:Pre?-true} or \cref{thm:Pre?-false} to obtain $V |-_{\Sigma,\Omega} t^\dir{d}$ or $V |/-_{\Sigma,\Omega} t^\dir{d}$.
\end{proof}

%We can also establish the converse of \cref{thm:Pre?-true,thm:Pre?-false}.
%
%\begin{corollary}\label{thm:toPre?-true}
%If\/ $V |-_{\Sigma,\Omega} t^\dir{d}$, then $V |-_{\Sigma,\Omega} t^{\dir{d}\,\BTrue\,s}$ for some boolean~$s$.
%\end{corollary}
%
%\begin{corollary}\label{thm:toPre?-false}
%If\/ $V |/-_{\Sigma,\Omega} t^\dir{d}$, then $V |-_{\Sigma,\Omega} t^{\dir{d}\,\BFalse\,s}$ for some boolean~$s$.
%\end{corollary}
%
%Whereas it is possible to prove \cref{thm:toPre?-true} directly by induction, converting a mode derivation to a generalised one, the negated antecedent of \cref{thm:toPre?-false} provides little information for constructing a required derivation, and basically we have to construct such a derivation from scratch for a raw term.
%But this is exactly what generalised mode preprocessing does, so we can simply reuse \cref{thm:generalised-mode-preprocessing}.
%And \cref{thm:toPre?-true} can also be proved easily in the same way.
%
%\begin{proof}[Proofs of \cref{thm:toPre?-true,thm:toPre?-false}]
%
%By \cref{thm:generalised-mode-preprocessing}, there is a derivation of $V |-_{\Sigma,\Omega} t^{\dir{d}\,g\,s}$ for some booleans $g$~and~$s$.
%If $V |-_{\Sigma,\Omega} t^\dir{d}$, then $g$~has to be $\BTrue$ since $g = \BFalse$ leads to a contradiction with \cref{thm:Pre?-false}.
%Symmetrically, if $V |/-_{\Sigma,\Omega} t^\dir{d}$, then $g$~has to be~$\BFalse$ so as not to contradict \cref{thm:Pre?-true}.
%\end{proof}

### 2023-07-09 tex/sec5-synthesis.tex
[ tex ] Use \nvdash instead of \not\vdash

    \item \SynRule{Anno}: For $\erase{\Gamma} |-_{\Sigma, \Omega} (\isTerm{t} \annotate A)^{\syn}$, it is decidable whether $\Gamma |- \isTerm{t} \chk A$ by induction hypothesis.
      \begin{itemize}
        \item If $\Gamma |- \isTerm{t} \chk A$, then $\Gamma |- \isTerm{t \annotate A} \syn A$.
        \item If $\Gamma \not|- t \chk A$ but $\Gamma |- \isTerm{t \annotate A} \syn$, then by inversion $\Gamma |- t \chk A$, leading to a contradiction.
      \end{itemize}

    \item \ChkRule{Sub}: If $\erase{\Gamma} |-_{\Sigma, \Omega}t^{\chk}$ by the rule $\ChkRule{Sub}$, then $\Gamma |-_{\Sigma, \Omega} t^{\syn}$ by inversion.
      We show that it is decidable for any $A$ whether $\Gamma |-_{\Sigma, \Omega} t \chk A$.
      By induction hypothesis, it is decidable whether $\Gamma |- t \syn B$ for some $B$:
      \begin{itemize}
        \item If $\Gamma \not|- t \syn C$ for any type $C$ but $\Gamma |- t \chk A$ is derivable, then by inversion $\Gamma |- t \syn B$ for some $B = A$, thus a contradiction.
        \item If $\Gamma |- t \syn B$ for some $B$, then either $A = B$ or $A \neq B$ by the decidable equality on $\Type_{\Sigma}(\Xi)$: 
          \begin{itemize}
            \item if $A = B$ then we have\/ $\Gamma |- t \chk A$;
            \item if $A \neq B$ but $\Gamma |- t \chk A$, then by inversion $\Gamma |- t \syn A$.
              However, by \Cref{thm:unique-syn}, synthesised types $A$ and $B$ must be equal, so we derive a contradiction.
          \end{itemize}
      \end{itemize}
    \item \Rule{Op}:
      Consider the mode derivation of\/ $\erase{\Gamma} |-_{\Sigma, \Omega} \tmOpts^{\dir{d}}$.
%
%  \begin{claim}\label{lem:args-induction}
%    For the argument list $\overrightarrow{[\Delta_i] A_{i}^{\dir{d_i}}}$, \emph{partial} substitution $\rho$ from $\Xi$ to $\emptyset$
%    \begin{enumerate}
%      \item either there is a minimal extension $\ext{\rho}$ of $\rho$ defined on $\synvar(\mathit{as})$ such that all of following judgements are derivable
%        \[
%          \Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\ext{\rho}} |-_{\Sigma, \Omega} \isTerm{t_i} \colon \simsub{A_i}{\ext{\rho}}^{\dir{d_i}}
%        \]
%
%      \item or there is no extension $\sigma$ of $\rho$ such that all $\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\sigma} |-_{\Sigma, \Omega} \isTerm{t_i} \colon \simsub{A_i}{\sigma}^{\dir{d_i}}$ has a typing derivation. 
%    \end{enumerate}
%  \end{claim}
  \end{itemize}
\end{proof}

@@

\begin{corollary}[Trichotomy on Raw Terms]\label{cor:trichotomy}
  For any mode-correct bidirectional type system $(\Sigma, \Omega)$, 
  exactly one of the following holds:
  \begin{enumerate}
    \item $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$ and\/ $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ for some type $A$.
    \item $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$ but\/ $\Gamma \not|-_{\Sigma, \Omega} \isTerm{t} : A$.
    \item $\erase{\Gamma} \not|-_{\Sigma, \Omega} \isTerm{t}^{\syn}$.
  \end{enumerate}
\end{corollary}
\begin{proof}
  Combine \Cref{lem:soundness-completeness,thm:mode-preprocessing} with \Cref{thm:bidirectional-type-synthesis-checking}.

>>>>

    \item \SynRule{Anno}: For $\erase{\Gamma} |-_{\Sigma, \Omega} (\isTerm{t} \annotate A)^{\syn}$, it is decidable whether $\Gamma |- \isTerm{t} \chk A$ by induction hypothesis.
      \begin{itemize}
        \item If $\Gamma |- \isTerm{t} \chk A$, then $\Gamma |- \isTerm{t \annotate A} \syn A$.
        \item If $\Gamma |/- t \chk A$ but $\Gamma |- \isTerm{t \annotate A} \syn$, then by inversion $\Gamma |- t \chk A$, leading to a contradiction.
      \end{itemize}

    \item \ChkRule{Sub}: If $\erase{\Gamma} |-_{\Sigma, \Omega}t^{\chk}$ by the rule $\ChkRule{Sub}$, then $\Gamma |-_{\Sigma, \Omega} t^{\syn}$ by inversion.
      We show that it is decidable for any $A$ whether $\Gamma |-_{\Sigma, \Omega} t \chk A$.
      By induction hypothesis, it is decidable whether $\Gamma |- t \syn B$ for some $B$:
      \begin{itemize}
        \item If $\Gamma |/- t \syn C$ for any type $C$ but $\Gamma |- t \chk A$ is derivable, then by inversion $\Gamma |- t \syn B$ for some $B = A$, thus a contradiction.
        \item If $\Gamma |- t \syn B$ for some $B$, then either $A = B$ or $A \neq B$ by the decidable equality on $\Type_{\Sigma}(\Xi)$: 
          \begin{itemize}
            \item if $A = B$ then we have\/ $\Gamma |- t \chk A$;
            \item if $A \neq B$ but $\Gamma |- t \chk A$, then by inversion $\Gamma |- t \syn A$.
              However, by \Cref{thm:unique-syn}, synthesised types $A$ and $B$ must be equal, so we derive a contradiction.
          \end{itemize}
      \end{itemize}
    \item \Rule{Op}:
      Consider the mode derivation of\/ $\erase{\Gamma} |-_{\Sigma, \Omega} \tmOpts^{\dir{d}}$.
%
%  \begin{claim}\label{lem:args-induction}
%    For the argument list $\overrightarrow{[\Delta_i] A_{i}^{\dir{d_i}}}$, \emph{partial} substitution $\rho$ from $\Xi$ to $\emptyset$
%    \begin{enumerate}
%      \item either there is a minimal extension $\ext{\rho}$ of $\rho$ defined on $\synvar(\mathit{as})$ such that all of following judgements are derivable
%        \[
%          \Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\ext{\rho}} |-_{\Sigma, \Omega} \isTerm{t_i} \colon \simsub{A_i}{\ext{\rho}}^{\dir{d_i}}
%        \]
%
%      \item or there is no extension $\sigma$ of $\rho$ such that all $\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\sigma} |-_{\Sigma, \Omega} \isTerm{t_i} \colon \simsub{A_i}{\sigma}^{\dir{d_i}}$ has a typing derivation. 
%    \end{enumerate}
%  \end{claim}
  \end{itemize}
\end{proof}

@@

\begin{corollary}[Trichotomy on Raw Terms]\label{cor:trichotomy}
  For any mode-correct bidirectional type system $(\Sigma, \Omega)$, 
  exactly one of the following holds:
  \begin{enumerate}
    \item $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$ and\/ $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ for some type $A$.
    \item $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$ but\/ $\Gamma |/-_{\Sigma, \Omega} \isTerm{t} : A$.
    \item $\erase{\Gamma} |/-_{\Sigma, \Omega} \isTerm{t}^{\syn}$.
  \end{enumerate}
\end{corollary}
\begin{proof}
  Combine \Cref{lem:soundness-completeness,thm:mode-preprocessing} with \Cref{thm:bidirectional-type-synthesis-checking}.
